### 2016-03-14 thesis/acknowledgement.tex
acknowl.

alister, duncan, c blake, greg

karl,

gonzalo, rob, piero, glenn, anna, angela,

elisa, shaun, silvia, james

gym friends, rock climbing friends

>>>>

First thanks go to my supervisor Alister, 
for offering me a PhD project in the country I always dreamt to live in, 
for guiding me during this long journey, 
for his patience and support, 
for all the lessons I've learnt, 
and for being constantly available to all of his students, 
no matters how many late nights he had to spend in the office correcting my writings. 
Second thanks go to my co-supervisor Duncan for the constructing scientific conversations that we had, 
for his useful comments and suggestions, 
and for his genuine interest in making sure that at any stage everything was going well. 
Very special thanks are reserved to Chris Blake, Duncan, Greg, and Karl, 
for their invaluable, immediate, unconditionate support, when I most than ever needed it.

%To my family and in particular to zia Anna and Suzy for being such enviable badasses 
%and such great examples. 
%
%To my PhD brothers and sisters, Angela, Anna, Gonzalo, Pierluigi, and Rob, 
%I'm glad I've been on this boat in your excellent company, guys. 
%Many thanks also to Glenn and Jeff for your informal and very comfortable attitude, 
%for always encouraging me to do my best and for all the good laughs we've had. 
%
%To my irreplaceable friends Elisa, Gonzalo, Greg $\heartsuit$ Anna, James, Shaun, Silvia, 
%thank you very much for being real. 
%
%And last, to Garry, because you might have glimpsed a tinder that already was inside me, 
%but you are doing a very good job at making it a blaze.

### 2016-03-14 thesis/conclusions.tex
acknowl.

In this thesis, we explored scaling relations between the supermassive black hole mass 
and various properties of the host spheroid, 
aiming at a more profound understanding of the co-evolution between SMBHs and galaxies. 
A summary of our principal findings and some additional considerations are presented here. 
This Chapter incorporates also some promising future research directions. \\

@@

we do not consider possible for current automatic routines to reproduce our analysis. \\

@@

In Chapter \ref{ch:mm} \citep{paperII}, we used our dataset to derive and explore 
the $M_{\rm BH} - L_{\rm gal}$ and $M_{\rm BH} - L_{\rm sph}$ (or $M_{\rm BH} - M_{\rm *,sph}$) diagrams. 
When considering all galaxies, irrespective of their morphological type, 
the $M_{\rm BH} - L_{\rm sph}$ correlation has a lower level of intrinsic scatter 
than the $M_{\rm BH} - L_{\rm gal}$ correlation. 
However, when considering only early-type galaxies (elliptical+lenticular), 
$L_{\rm gal}$ and $L_{\rm sph}$ correlate equally well with $M_{\rm BH}$. 
\citet{lasker2014anal} found the same level of intrinsic scatter in their 
$M_{\rm BH} - L_{\rm gal}$ and $M_{\rm BH} - L_{\rm sph}$ diagrams for all galaxies
because their sample was highly biased towards early-type objects 
(among their 35 galaxies, only four are spirals). 
Our subsample of S\'ersic spheroids defines a less steep $M_{\rm BH} - L_{\rm sph}$ 
sequence than previously reported by \citet{grahamscott2013}. 
We did not observe any offset of spheroids with S\'ersic index $n_{\rm sph} < 2$ 
from the $M_{\rm BH} - L_{\rm sph}$ correlation traced by spheroids with $n_{\rm sph} > 2$, 
in disagreement with the claims of \citet{sani2011}. 
We identified two distinct trends in the $M_{\rm BH} - M_{\rm *,sph}$ diagram, 
that is, a \emph{red sequence} of early-type galaxies (elliptical+lenticular) 
following $M_{\rm BH} \propto M_{\rm *,sph}^{1.04 \pm 0.10}$ 
and a dramatically steeper \emph{blue sequence} of late-type galaxies (spiral) 
following $M_{\rm BH} \propto M_{\rm *,sph}^{2-3}$. 
This is in agreement with the near-linear $M_{\rm BH} - L_{\rm sph}$ (or $M_{\rm BH} - M_{\rm *,sph}$) correlation 
measured by previous studies that used galaxy samples dominated by high-mass early-type objects 
(e.g.~\citealt{magorrian1998,marconihunt2003,haringrix2004,gultelkin2009,sani2011,beifiori2012,
erwingadotti2012,vika2012}),  
and it gives reason as to why other studies, 
whose galaxy samples gave a better representation of the late-type population 
with low-mass ($M_{\rm BH} \lesssim 10^8 \rm~M_\odot$) black holes, 
obtained a correlation steeper than linear 
or noticed super-linear deviations at the low-luminosity end of the diagram 
(e.g.~\citealt{laor1998,laor2001,wandel1999,salucci2000,ryan2007}). 
An interesting exception has been the under-linear $M_{\rm BH} \propto L_{\rm sph}^{0.75 \pm 0.10}$ relation 
measured by \citet{lasker2014anal}, 
who attributed the smaller log-slope of the correlation 
to the reduction of bulge luminosities derived from their multicomponents fits 
and to the type of linear regression algorithm used. 
It is worth mentioning that, among our 17 spiral galaxies, at least 12 host a Seyfert nucleus. 
\citet{wandel1999} and \citet{ryan2007} had previously noted that a linear $M_{\rm BH} - M_{\rm *,sph}$ relation 
systematically overestimates the $M_{\rm BH}/M_{\rm *,sph}$ ratio in Seyfert galaxies. 
In effect, the sample of $\approx 140$ low-redshift AGNs with virial black hole masses 
$10^5 \lesssim M_{\rm BH}/{\rm M_\odot} \lesssim 2 \times 10^6$ collected by \citet{grahamscott2015} 
appear to be the low-mass continuation of the $M_{\rm BH} - M_{\rm *,sph}$ sequence traced by our spiral galaxies. 
Finally, we note that our estimate of the normalization of the $M_{\rm BH} - M_{\rm *,sph}$ relation for early-type galaxies 
is the largest ever reported: $M_{\rm BH}/M_{\rm *,sph}$ is $\approx 0.7\%$ using our dataset, 
whereas it is as low as $\approx 0.5\%$ in \citet{scott2013} and \citet{kormendyho2013}, 
and even lower in previous studies. 
A recalibration of the $M_{\rm BH} - M_{\rm *,sph}$ normalization has important consequences 
for semi-analytical models and simulations of galaxy evolution that include black hole growth, 
and it also affects calculations of gravitational waves background from SMBH binary coalescence 
(but see \citealt{shankar2016}). \\

m-n? \\

@@

%(but see Shankar et al.~\emph{submitted to MNRAS} TBD). \\

>>>>

In this thesis, we explored scaling relations between the supermassive black hole mass 
and various properties of the host spheroid, 
aiming at a more profound understanding of the co-evolution between SMBHs and their host galaxies. 
A summary of our principal findings and some additional considerations are presented here. 
This Chapter incorporates also some promising future research directions. \\

@@

we do not consider possible for current automatic routines to reproduce our human-supervised analysis. \\

@@

In Chapter \ref{ch:mm} \citep{paperII}, we used our dataset to derive and explore 
the $M_{\rm BH} - L_{\rm gal}$ and $M_{\rm BH} - L_{\rm sph}$ (or $M_{\rm BH} - M_{\rm *,sph}$) diagrams. 
When considering all galaxies, irrespective of their morphological type, 
the $M_{\rm BH} - L_{\rm sph}$ correlation has a lower level of intrinsic scatter 
than the $M_{\rm BH} - L_{\rm gal}$ correlation. 
However, when considering only early-type galaxies (elliptical+lenticular), 
$L_{\rm gal}$ and $L_{\rm sph}$ correlate equally well with $M_{\rm BH}$. 
\citet{lasker2014anal} found the same level of intrinsic scatter in their 
$M_{\rm BH} - L_{\rm gal}$ and $M_{\rm BH} - L_{\rm sph}$ diagrams for all galaxies
because their sample was highly biased towards early-type objects 
(among their 35 galaxies, only four are spirals). 
Our subsample of S\'ersic spheroids defines a less steep $M_{\rm BH} - L_{\rm sph}$ 
sequence than previously reported by \citet{grahamscott2013}. 
We did not observe any offset of spheroids with S\'ersic index $n_{\rm sph} < 2$ 
from the $M_{\rm BH} - L_{\rm sph}$ correlation traced by spheroids with $n_{\rm sph} > 2$, 
in disagreement with the claims of \citet{sani2011}. 
We identified two distinct trends in the $M_{\rm BH} - M_{\rm *,sph}$ diagram, 
that is, a \emph{red sequence} of early-type galaxies (elliptical+lenticular) 
following $M_{\rm BH} \propto M_{\rm *,sph}^{1.04 \pm 0.10}$ 
and a dramatically steeper \emph{blue sequence} of late-type galaxies (spiral) 
following $M_{\rm BH} \propto M_{\rm *,sph}^{2-3}$. 
This is in agreement with the near-linear $M_{\rm BH} - L_{\rm sph}$ (or $M_{\rm BH} - M_{\rm *,sph}$) correlation 
measured by previous studies that used galaxy samples dominated by high-mass early-type objects 
(e.g.~\citealt{magorrian1998,marconihunt2003,haringrix2004,gultelkin2009,sani2011,beifiori2012,
erwingadotti2012,vika2012}),  
and it gives reason as to why other studies, 
whose galaxy samples gave a better representation of the late-type population 
with low-mass ($M_{\rm BH} \lesssim 10^8 \rm~M_\odot$) black holes, 
obtained a correlation steeper than linear 
or noticed super-linear deviations at the low-luminosity end of the diagram 
(e.g.~\citealt{laor1998,laor2001,wandel1999,salucci2000,ryan2007}). 
An interesting exception has been the under-linear $M_{\rm BH} \propto L_{\rm sph}^{0.75 \pm 0.10}$ relation 
measured by \citet{lasker2014anal}, 
who attributed the smaller log-slope of the correlation 
to the reduction of bulge luminosities derived from their multicomponents fits 
and to the type of linear regression algorithm used. 
It is worth mentioning that, among our 17 spiral galaxies, at least 12 host a Seyfert nucleus. 
\citet{wandel1999} and \citet{ryan2007} had previously noted that a linear $M_{\rm BH} - M_{\rm *,sph}$ relation 
systematically overestimates the $M_{\rm BH}/M_{\rm *,sph}$ ratio in Seyfert galaxies. 
In effect, the sample of $\approx 140$ low-redshift AGNs with virial black hole masses 
$10^5 \lesssim M_{\rm BH}/{\rm M_\odot} \lesssim 2 \times 10^6$ collected by \citet{grahamscott2015} 
appear to be the low-mass continuation of the $M_{\rm BH} - M_{\rm *,sph}$ sequence traced by our spiral galaxies. 
Finally, we note that our estimate of the normalization of the $M_{\rm BH} - M_{\rm *,sph}$ relation for early-type galaxies 
is the largest ever reported: $M_{\rm BH}/M_{\rm *,sph}$ is $\approx 0.7\%$ using our dataset, 
whereas it is as low as $\approx 0.5\%$ in \citet{scott2013} and \citet{kormendyho2013}, 
and even lower in previous studies. 
A recalibration of the $M_{\rm BH} - M_{\rm *,sph}$ normalization has important consequences 
for semi-analytical models and simulations of galaxy evolution that include black hole growth, 
and it also affects calculations of gravitational waves background from SMBH binary coalescence. 
However, it is important to keep in mind the warnings of \citeauthor{merritt2013book} 
(\citeyear{merritt2013book}; see also \citealt{ferraresemerritt2000,merrittferrarese2001ms,merrittferrarese2001sphinfl,
valluri2004,ferrareseford2005}), 
who cautioned that the majority of stellar-dynamics based black hole mass measurements 
are likely to have been overestimated by a factor of $3$ to $4$ 
due to an insufficient spatial resolution of the sphere-of-influence of black holes. 
His concern seems to be confirmed by the very recent findings of \citet{shankar2016}, 
who argued that the current sample of directly measured black hole masses 
is highly biased by the fact that their detectability 
is subject to the requirement of spatially resolve their gravitational sphere-of-influence. 
This selection effect allows us to see only the upper envelope of the ``true'' black hole mass correlations, 
artificially increasing the normalization of the $M_{\rm BH} - \sigma_*$ correlation by a factor of $\approx 3$ 
and even more dramatically that of the $M_{\rm BH} - M_{\rm *,sph}$ correlation. 
New black hole mass measurements obtained with the extraordinary spatial resolution of the next generation 
30-meter class telescopes will help shed light on this point.  \\

In Chapter \ref{ch:mn} \citep{paperIII}, 
we analysed the $L_{\rm sph} - n_{\rm sph}$ and $M_{\rm BH} - n_{\rm sph}$ diagrams 
and found that in the former early- and late-type galaxies split into two separate correlations, 
whereas in the latter all galaxies -- irrespective of their morphological type -- define a single correlation. 
Our $M_{\rm BH} - L_{\rm sph}$ and $M_{\rm BH} - n_{\rm sph}$ diagrams 
show consistent amount of intrinsic scatter, 
in agreement with the results from \citet{grahamdriver2007}. 
The present dataset allows one to explore also the relation of the black hole mass with 
the spheroid central surface density, $\mu_{\rm 0, sph}$, 
and the spheroid deprojected central stellar density, $\rho_{\rm *,0,sph}$. \\

@@

\vspace{5cm}
\begin{flushright}
\emph{Keep looking up. }
\end{flushright}
